### 2014-06-18 ApTranscriptome_TR.Rmd
Accidental but correct merge of rollback branch to master

**June 12, 2014**

@@

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
#biocLite("Rgraphviz")
#biocLite("RDAVIDWebService")
#biocLite("GOSemSim")
library(topGO) 
library(Rgraphviz)	
library(GOSemSim)

@@

```{r download, echo = TRUE}
### Annotation file

# from either AWS or GoogleDrive
annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
#a2 <- getURL("https://googledrive.com/host/0B75IymziRJ_9Tlg1U1Vxbjk1bzg") # GoogleDrive link

annotation.file <- read.csv(textConnection(annotation.URL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
str(annotation.file)

# Convert to data.table
annotation.table <- data.table(annotation.file)
head(annotation.table)
```

@@

# remove from TPM.dt.sub
setkey(TPM.dt.sub, Transcript)
TPM.dt.sub <- TPM.dt.sub[!cont.list]
str(TPM.dt.sub)
length(unique(TPM.dt.sub$Transcript))
```

@@

# plot
mosaic.colors <- brewer.pal(5, "Blues")

par(mar = c(5.1, 4.1, 4.1, 5.1))
mosaicplot(tt2, ylab = "A. picea", xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("High", "Low", "Bimodal", "Intermediate", "NotResp")
text(c(.1,.3,.55,.75,.95), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE)
text(par("usr")[2] + 0.25, c(.95, .75, .55, .35, .1), srt = 0, adj = 1, labels = labels, xpd = TRUE)
text(.5, par("usr")[4] + .1, "A. carolinensis", xpd = TRUE)
```

@@

I perform gene set enrichment analysis below, but a quick `grep` shows that there are 26 transcripts with GO term "response to stress", though this is not enriched compared to the frequency of this term in the full dataset.

@@

@@

@@

```{r geneid2go, echo=TRUE, eval=TRUE, cache=TRUE}
# create geneid2go.map file from FastAnnotator annotationtable.txt
mapname = "geneid2go.map"
if(file.exists(mapname)) print("Gene ID to GO term file exists") else{
    geneid2GOmap(annotation.file, ontology = c("BP", "CC", "MF"), mapname = mapname)
}
```

@@

```{r readmap, echo=TRUE, eval=TRUE}
# read mappings file
geneID2GO <- readMappings(file = "geneid2go.map")
str(head(geneID2GO))
```

### GSEA for thermally-responsive transcripts ###

@@

# Function to select top genes (defined above)
selectFDR <- function(padj) {
    return(padj < 0.05)
}

@@

# report items in each cluster at height 1.5
GSEAReportClusters(Ap.low.hclust, h = 1.5)
```

```{r gsea_Ap_bim, resuts = 'hide'}
## Genes with *Bimodal* expression in both colonies
Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ap.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.bim) <- names(Ap.geneList)
Ap.geneList.bim[(which(names(Ap.geneList.bim) %in% Ap.bim))] <- 1
# check correct number of values set to 1
table(Ap.geneList.bim)
# Run GSEA
Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
```

@@

Next, I perform GSEA for genes in each functional type in one species but not the other (e.g. the set difference) to gain insight on differences between the colonies.

```{r extract_Transcripts_only}
A22.only.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
Ar.only.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]

A22.only.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
Ar.only.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]

A22.only.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
Ar.only.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]

A22.only.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
Ar.only.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
```

```{r gsea_by_category}
# A22 'High' genes
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high.transcripts))] <- 1
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
```

```{r gsea_setdiff, results = 'hide'}
# A22 'High' genes not in Ar**
A22.geneList.high.only <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high.only) <- names(Ap.geneList)
A22.geneList.high.only[(which(names(A22.geneList.high.only) %in% A22.only.high))] <- 1
A22.high.only.gsea <- gsea(genelist = A22.geneList.high.only, geneID2GO = geneID2GO)

# *Ar 'High' genes not in A22**
Ar.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.high) <- names(Ap.geneList)
Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.only.high))] <- 1
Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)

# A22 'Low' genes not in Ar**
A22.geneList.low <- rep(0, length = length(Ap.geneList))
names(A22.geneList.low) <- names(Ap.geneList)
A22.geneList.low[(which(names(A22.geneList.low) %in% A22.only.low))] <- 1
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
# Ar 'Low' genes not in A22**
Ar.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.low) <- names(Ap.geneList)
Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.only.low))] <- 1
Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)

# A22 'Bimodal' genes not in Ar**
A22.geneList.bim <- rep(0, length = length(Ap.geneList))
names(A22.geneList.bim) <- names(Ap.geneList)
A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.only.bim))] <- 1
A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)

# Ar 'Bimodal' genes not in A22**
Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.bim) <- names(Ap.geneList)
Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.only.bim))] <- 1
Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)

# A22 'Intermediate' genes not in Ar**
A22.geneList.int <- rep(0, length = length(Ap.geneList))
names(A22.geneList.int) <- names(Ap.geneList)
A22.geneList.int[(which(names(A22.geneList.int) %in% A22.only.int))] <- 1
table(A22.geneList.int)
A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)

# Ar 'Intermediate' genes not in A22**
Ar.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.int) <- names(Ap.geneList)
Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.only.int))] <- 1
Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)

# combine into single table
A22.high.gsea$Type <- "High"
A22.low.gsea$Type <- "Low"
A22.int.gsea$Type <- "Intermediate"
A22.bim.gsea$Type <- "Bimodal"
A22.gsea <- rbind(A22.high.gsea, A22.low.gsea, A22.int.gsea, A22.bim.gsea)
A22.gsea$Species <- "ApVT"

@@

```{r gsea_setdiff_table, echo = FALSE, results = 'asis'}
#pandoc.table(Ap.setdiff.gsea[ , 1:4], caption = "Enriched GO terms found in one species but not the other for each thermal-response category")
print(xtable(Ap.setdiff.gsea[ , 1:4], comment = getOption("xtable.comment", FALSE)))
```

@@

```{r plot_GOstress, echo=FALSE, eval=TRUE}
high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]

>>>>

**June 16, 2014**

@@

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
#biocLite("Rgraphviz")
#biocLite("GOSemSim")
library(topGO) 
library(Rgraphviz)	
library(GOSemSim)

@@

```{r download, echo = TRUE}
# URL for annotation file
annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
# load 
annotation.file <- read.csv(textConnection(annotation.URL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
str(annotation.file)

# Convert to data.table
annotation.table <- data.table(annotation.file)
str(annotation.table)
```

@@

# remove from TPM.dt.sub
setkey(TPM.dt.sub, Transcript)
TPM.dt.sub <- TPM.dt.sub[!cont.list]
str(TPM.dt.sub)
length(unique(TPM.dt.sub$Transcript))

# remove from annotation.table
setkey(annotation.table, Sequence.Name)
annotation.table <- annotation.table[!cont.list]
str(annotation.table)
```

@@

# plot
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("High", "Low", "Bimodal", "Intermediate", "NotResp")
text(c(.1,.3,.55,.75,.95), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4], "A. carolinensis", xpd = TRUE, font = 3)
legend(1.05, .75, fill = mosaic.colors, c("High", "Low", "Bimodal", "Intermediate", "NotResp"), xpd=TRUE)
```

@@

### Proportion of responsive transcripts annotated

Of the responsive transcripts, `r prettyNum(length(which(responsive.lms.ann.type$best.hit.to.nr == "-"))/length(responsive.lms.ann.type$best.hit.to.nr)*100)`% are annotated, while `r prettyNum(length(which(annotation.table$best.hit.to.nr == "-"))/nrow(annotation.table) * 100)`% of all transcripts are annotated.

### Candidate gene enrichment

First, I explore of candidate genes with the GO term "response to stress" are enriched in the responsive data set.

@@

While `r nrow(GO0006950.responsive)` "response to stress" genes are in the responsive set, this is not enriched compared to the whole transcriptome.

@@

### Gene set enrichment analysis

@@

```{r geneid2go, echo=TRUE, eval=TRUE, cache=TRUE}
# create geneid2go.map file from FastAnnotator annotationtable.txt
mapname = "results/geneid2go.map"
if(file.exists(mapname)) print("Gene ID to GO term file exists") else{
    geneid2GOmap(as.data.frame(annotation.table), ontology = c("BP", "CC", "MF"), mapname = mapname)
}
```

@@

```{r readmap, echo=TRUE, eval=TRUE}
# read mappings file
geneID2GO <- readMappings(file = mapname)
str(head(geneID2GO))
```

### GSEA for thermally-responsive transcripts

@@

# Function to select genes with adjusted P-value < 0.05
selectFDR <- function(padj) {
    return(padj < 0.05)
}

@@

# report items in each cluster at height 2
GSEAReportClusters(Ap.low.hclust, h = 2)
```

```{r gsea_Ap_bim, resuts = 'hide'}
# Genes with *Bimodal* expression in both colonies
Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ap.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.bim) <- names(Ap.geneList)
Ap.geneList.bim[(which(names(Ap.geneList.bim) %in% Ap.bim))] <- 1
# check correct number of values set to 1
table(Ap.geneList.bim)
# Run GSEA
Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
```

@@

```{r gsea_Ar_int_A22_bim, results = 'hide'}
# genes with *Intermediate* expression in Ar and *Bimodal* in A22
Ar.int.A22.bim <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" & Ap.response.type$A22.type == "Bimodal"), "Transcript"]
Ar.int.A22.bim.geneList <- rep(0, length = length(Ap.geneList))
names(Ar.int.A22.bim.geneList) <- names(Ap.geneList)
Ar.int.A22.bim.geneList[(which(names(Ar.int.A22.bim.geneList) %in% Ar.int.A22.bim))] <- 1
# check correct number of values set to 1
table(Ar.int.A22.bim.geneList)
# run GSEA
Ar.int.A22.bim.gsea <- gsea(genelist = Ar.int.A22.bim.geneList, geneID2GO = geneID2GO, plotpath = NA)
```

```{r gsea_Ar_int_A22_bim_cluster}
# similarity among enriched GO terms
Ar.int.A22.bim.gsea.term.sim <- termSim(Ar.int.A22.bim.gsea$GO.ID, Ar.int.A22.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ar.int.A22.bim.gsea.term.sim) <- Ar.int.A22.bim.gsea$Term
colnames(Ar.int.A22.bim.gsea.term.sim) <- Ar.int.A22.bim.gsea$Term
# distance matrix
Ar.int.A22.bim.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.bim.gsea.term.sim))
plot(Ar.int.A22.bim.gsea.term.sim.hclust)

# report items in each cluster at height 1.5
GSEAReportClusters(Ar.int.A22.bim.gsea.term.sim.hclust, h = 1)
```

```{r gsea_Ar_int_A22_low, results = 'hide'}
# genes with *Intermediate* expression in Ar and *Low* in A22
Ar.int.A22.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" & Ap.response.type$A22.type == "Low"), "Transcript"]
Ar.int.A22.low.geneList <- rep(0, length = length(Ap.geneList))
names(Ar.int.A22.low.geneList) <- names(Ap.geneList)
Ar.int.A22.low.geneList[(which(names(Ar.int.A22.low.geneList) %in% Ar.int.A22.low))] <- 1
# check correct number of values set to 1
table(Ar.int.A22.low.geneList)
# run GSEA
Ar.int.A22.low.gsea <- gsea(genelist = Ar.int.A22.low.geneList, geneID2GO = geneID2GO, plotpath = NA)
```

```{r gsea_Ar_int_A22_low_cluster}
# similarity among enriched GO terms
Ar.int.A22.low.gsea.term.sim <- termSim(Ar.int.A22.low.gsea$GO.ID, Ar.int.A22.low.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ar.int.A22.low.gsea.term.sim) <- Ar.int.A22.low.gsea$Term
colnames(Ar.int.A22.low.gsea.term.sim) <- Ar.int.A22.low.gsea$Term
# distance matrix
Ar.int.A22.low.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.low.gsea.term.sim))
plot(Ar.int.A22.low.gsea.term.sim.hclust)
# report items in each cluster at height 1.8
GSEAReportClusters(Ar.int.A22.low.gsea.term.sim.hclust, h = 1.6)
```

### GSEA for each functional type in each species

GO enrichment for *A. picea* **High** genes.

```{r gsea_A22_high, results = 'hide'}
# A22 'High' genes
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high.transcripts))] <- 1
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. picea* **High** genes.

```{r gsea_A22_high_cluster}
# similarity among enriched GO terms
A22.high.gsea.term.sim <- termSim(A22.high.gsea$GO.ID, A22.high.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(A22.high.gsea.term.sim) <- A22.high.gsea$Term
colnames(A22.high.gsea.term.sim) <- A22.high.gsea$Term
# distance matrix
A22.high.gsea.term.sim.hclust <- hclust(dist((A22.high.gsea.term.sim)))
plot((A22.high.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(A22.high.gsea.term.sim.hclust, h = 1.2)
```
GO enrichment for *A. picea* **Low** genes.

```{r gsea_A22_low, results = 'hide'}
# A22 'Low' genes
A22.geneList.low <- rep(0, length = length(Ap.geneList))
names(A22.geneList.low) <- names(Ap.geneList)
A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low.transcripts))] <- 1
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. picea* **Low** genes.

```{r gsea_A22_low_cluster}
# similarity among enriched GO terms
A22.low.gsea.term.sim <- termSim(A22.low.gsea$GO.ID, A22.low.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(A22.low.gsea.term.sim) <- A22.low.gsea$Term
colnames(A22.low.gsea.term.sim) <- A22.low.gsea$Term
# distance matrix
A22.low.gsea.term.sim.hclust <- hclust(dist((A22.low.gsea.term.sim)))
plot((A22.low.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(A22.low.gsea.term.sim.hclust, h = 2)
```

GO enrichment for *A. picea* **Bimodal** genes.

```{r gsea_A22_bim, results = 'hide'}
# A22 'Bim' genes
A22.geneList.bim <- rep(0, length = length(Ap.geneList))
names(A22.geneList.bim) <- names(Ap.geneList)
A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.bim.transcripts))] <- 1
A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. picea* **Low** genes.

```{r gsea_A22_bim_cluster}
# similarity among enriched GO terms
A22.bim.gsea.term.sim <- termSim(A22.bim.gsea$GO.ID, A22.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(A22.bim.gsea.term.sim) <- A22.bim.gsea$Term
colnames(A22.bim.gsea.term.sim) <- A22.bim.gsea$Term
# distance matrix
A22.bim.gsea.term.sim.hclust <- hclust(dist((A22.bim.gsea.term.sim)))
plot((A22.bim.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(A22.bim.gsea.term.sim.hclust, h = 1)
```

GO enrichment for *A. picea* **Intermediate** genes.

```{r gsea_A22_int, results = 'hide'}
# A22 'Int' genes
A22.geneList.int <- rep(0, length = length(Ap.geneList))
names(A22.geneList.int) <- names(Ap.geneList)
A22.geneList.int[(which(names(A22.geneList.int) %in% A22.int.transcripts))] <- 1
A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. picea* **Intermediate** genes.

```{r gsea_A22_int_cluster}
# similarity among enriched GO terms
A22.int.gsea.term.sim <- termSim(A22.int.gsea$GO.ID, A22.int.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(A22.int.gsea.term.sim) <- A22.int.gsea$Term
colnames(A22.int.gsea.term.sim) <- A22.int.gsea$Term
# distance matrix
A22.int.gsea.term.sim.hclust <- hclust(dist((A22.int.gsea.term.sim)))
plot((A22.int.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(A22.int.gsea.term.sim.hclust, h = 1.3)
```

```{r gsea_combine, eval = FALSE}
# combine into single table
A22.high.gsea$Type <- "High"
A22.low.gsea$Type <- "Low"
A22.int.gsea$Type <- "Intermediate"
A22.bim.gsea$Type <- "Bimodal"
A22.gsea <- rbind(A22.high.gsea, A22.low.gsea, A22.int.gsea, A22.bim.gsea)
A22.gsea$Species <- "ApVT"

@@

```{r gsea_setdiff_table, echo = FALSE, results = 'asis', eval = FALSE}
#pandoc.table(Ap.setdiff.gsea[ , 1:4], caption = "Enriched GO terms found in one species but not the other for each thermal-response category")
print(xtable(Ap.setdiff.gsea[ , 1:4], comment = getOption("xtable.comment", FALSE)))
```

@@

```{r plot_GOstress, echo=FALSE, eval=FALSE}
high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]
